---
title: "Multi-Dimensional Dynamic Systems"
author: "Andre Bourbonnais"
date: "2023-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 2, Dynamic systems

Multidimensional ablalalalala

```{r, echo = T, message = F, warning = F}
library(tidyverse)
library(deSolve)
library(ggpubr)
```

## 1. Lotka-Volterra predator-prey equations

Lotka-Volterra (LV) predator-prey equations are given by the following system of differential equations:

$$\Large \frac{dn}{dt} = rn-anp$$ $$\Large\frac{dp}{dt} = anp - \mu p$$

Where $\large n(t)$ is the prey population, $\large p(t)$ is the predator population, $\large r$ is the intrisic prey growth rate, $\large a$ is the predation rate (*attack rate*), and $\large \mu$ is the predator death rate (*mortality*).

### a) Write the *isoclines* of the system.

*"What will the **predator** population be when the **prey** population growth equals zero?"* Meaning, there is a sweet spot in predator population where the prey population is stable. This is called the *prey isocline* and *vice versa*. This value is called the *isocline* because it is the line where the population is stable.

Thus, solve each *differential equation* at a time (equal to zero) and we will get the corresponding population value to keep the other value at 0 growth.

$$\large \frac{dn}{dt} = 0 \leftrightarrow rn-anp = 0 \leftrightarrow n(r-ap) = 0$$ $$\large \frac{dp}{dt} = 0 \leftrightarrow anp - \mu p = 0 \leftrightarrow p(an-\mu) = 0$$\
We get two trivial equilibrium points, $\large n = 0, p=0$ and two equilibrium points of interest: $$\Large \text{Prey: } p = \frac{r}{a}$$ $$\Large \text{Predator: } n = \frac{\mu}{a}$$

```{r}
# Define the function for LV isoclines
LV_iso <- function(r, a, mu){
  ne <- r/a
  pe <- mu/a
  return(list(prey_iso = ne, pred_iso = pe))
}

# Define the parameters
P <- list(r = 3,
          a = 0.2,
          mu = 0.6
          )
isoc <- LV_iso(P$r, P$a, P$mu)

# Plot the isoclines

# Create a blank ggplot object with the axes limits where i add the isoclines 
# using horizontal and vertical lines.
ggplot() + 
  geom_hline(yintercept = isoc$prey_iso, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = isoc$pred_iso, color = "red", linetype = "dashed") +
  labs(x = "Prey", y = "Predator") +
  # Multiply by 2 as the isoclines are the equilibrium points.
  xlim(0, isoc$pred_iso*2) + 
  ylim(0, isoc$prey_iso*2) +
  theme_bw() +
  annotate("text", x = 5, y = 16, label = "Prey isocline", color = "blue") +
  annotate("text", x = 3.2, y = 24, label = "Predator isocline", color = "red",
           angle = -90) +
  ggtitle("Lotka-Volterra Isoclines")

```

## s
