---
title: "Multi-Dimensional Dynamic Systems"
author: "Andre Bourbonnais"
date: "2023-11-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 2, Dynamic systems

Multidimensional ablalalalala

```{r, echo = T, message = F, warning = F}
library(tidyverse)
library(deSolve)
library(ggpubr)
```

## 1. Lotka-Volterra predator-prey equations

Lotka-Volterra (LV) predator-prey equations are given by the following system of differential equations:

$$\Large \frac{dn}{dt} = rn-anp$$ $$\Large\frac{dp}{dt} = anp - \mu p$$

Where $\large n(t)$ is the prey population, $\large p(t)$ is the predator population, $\large r$ is the intrisic prey growth rate, $\large a$ is the predation rate (*attack rate*), and $\large \mu$ is the predator death rate (*mortality*).

### a) Write the *isoclines* of the system.

*"What will the **predator** population be when the **prey** population growth equals zero?"* Meaning, there is a sweet spot in predator population where the prey population is stable. This is called the *prey isocline* and *vice versa*. This value is called the *isocline* because it is the line where the population is stable.

Thus, solve each *differential equation* at a time (equal to zero) and we will get the corresponding population value to keep the other value at 0 growth.

$$\large \frac{dn}{dt} = 0 \leftrightarrow rn-anp = 0 \leftrightarrow n(r-ap) = 0$$ $$\large \frac{dp}{dt} = 0 \leftrightarrow anp - \mu p = 0 \leftrightarrow p(an-\mu) = 0$$\
We get two trivial equilibrium points, $\large n = 0, p=0$ and two equilibrium points of interest: $$\Large \text{Prey: } p = \frac{r}{a}, \quad \text{Predator: } n = \frac{\mu}{a}$$

```{r}
# Define the function for LV isoclines
LV_iso <- function(r, a, mu){
  ne <- r/a
  pe <- mu/a
  return(list(prey_iso = ne, pred_iso = pe))
}

# Define the parameters
P <- list(r = 3,
          a = 0.2,
          mu = 0.6
          )
isoc <- LV_iso(P$r, P$a, P$mu)

# Plot the isoclines

# Create a blank ggplot object with the axes limits where i add the isoclines 
# using horizontal and vertical lines.
ggplot() + 
  geom_hline(yintercept = isoc$prey_iso, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = isoc$pred_iso, color = "red", linetype = "dashed") +
  labs(x = "Prey", y = "Predator") +
  # Multiply by 2 as the isoclines are the equilibrium points.
  xlim(0, isoc$pred_iso*2) + 
  ylim(0, isoc$prey_iso*2) +
  theme_bw() +
  annotate("text", x = 5, y = 16, label = "Prey isocline", color = "blue") +
  annotate("text", x = 3.2, y = 24, label = "Predator isocline", color = "red",
           angle = -90) +
  ggtitle("Lotka-Volterra Isoclines")

```

### b) Simulate using `ode()` and plot a time-plot and phase-plane plot.

Run simulations with ode() from deSolve in R. Use two state variables: prey and predator densities. Return their time derivatives as a vector in a list. Simulate the system, outputting a matrix with time, prey density, and predator density. Plot both populations over time and create a phase-plane plot (predator vs. prey). Incorporate isoclines using LV_iso().

1.  **Simulation Setup**: Utilize the `ode()` function from the `deSolve` package.

2.  **State Variables**: Focus on two variables - prey and predator densities.

3.  **Output Format**: System function should return a list with a vector of these time derivatives.

4.  **Run Simulation**: Generate a matrix output with columns for time, prey density, and predator density.

5.  **Plotting**: - Create a time plot showing both populations. - Construct a phase-plane plot with predator vs. prey densities.

6.  **Isoclines**: Use `LV_iso()` function to add isoclines to the phase-plane plot.

```{r}
# Define the function for LV isoclines
LV_iso <- function(r, a, mu){
  ne <- r/a
  pe <- mu/a
  return(list(prey_iso = ne, pred_iso = pe))
}


# LV system function
LV_sys <- function(t, np, P){
  # State variables
  n <- np[1]
  p <- np[2]
  
  # Differential equations
  dndt <- P$r*n - P$a*n*p
  dpdt <- P$a*n*p - P$mu*p
  
  return(list(c(dndt, dpdt)))
}

# Define the parameters
P2 <- list(
  r = 2,
  a = 1,
  mu = 2
  )

# Define the initial population size
np0 <- c(n = 2, p = 1)

# Time range
time_vec <- seq(0, 10, 0.1)

# Run the simulation
res_sim <- ode(y = np0, times = time_vec, func = LV_sys, parms = P2)

# Pivot the data to a long formated dataframe
df_sim <- res_sim %>%
  as.data.frame() %>%
  pivot_longer(cols = c("n", "p"), names_to = "Species",
               values_to = "Population")

# Calculate the isoclines
isoc <- LV_iso(P2$r, P2$a, P2$mu)

# Plotting the results
timeplot <- df_sim %>% 
  ggplot(aes(x = time, y = Population, color = Species)) +
  geom_line() +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw() +
  labs(x = "Time", y = "Population") +
  ggtitle("Lotka-Volterra Predator-Prey Simulation") 

# Phase plane plot
phase_plane <- df_sim %>% 
  pivot_wider(names_from = Species, values_from = Population) %>%
  ggplot(aes(x = n, y = p)) +
  geom_path() +
  geom_hline(yintercept = isoc$prey_iso, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = isoc$pred_iso, color = "red", linetype = "dashed") +
  labs(x = "Prey", y = "Predator") +
  # Multiply by 2 as the isoclines are the equilibrium points.
  xlim(0, isoc$pred_iso*2) + 
  ylim(0, isoc$prey_iso*2) +
  theme_bw() +
  annotate("text", x = isoc$pred_iso+1, y = isoc$prey_iso+0.1,
           label = "Prey isocline", color = "blue", size = 2) +
  annotate("text", x = isoc$pred_iso+0.05, y = isoc$prey_iso+0.5,
           label = "Predator isocline", color = "red", angle = -90, size = 2) +
  ggtitle("Lotka-Volterra Phase Plane Plot")

ggarrange(timeplot, phase_plane, ncol = 1, nrow = 2)

```

### c) Calculate the Jacobian matrix

To determine the stability of an equilibrium, one should calculate the *Jacobian* matrix of the system at the equilibrium point. If all *eigenvalues* of the Jacobian matrix have a **negative real part**, the equilibrium is stable.\

Calculate the Jacobian matrix for the system at the equilibrium point. Write a function that takes three parameters ($\large r,a,\mu$) as input and returns the Jacobian matrix as output. We have extended the LV equation with $\large c$ as the conversion efficiency of prey to predator biomass:

$$\Large\frac{dn}{dt} = f_n(n,p) =rn-anp$$ $$\Large\frac{dp}{dt} = f_p(n,p)=canp - \mu p$$ **i) Calculating the Jacobian matrix:** The partial derivative of $f_n$ with respect to $n$ is written $\large \frac{\partial f_n}{\partial n}$. It can be interpreted as the slope of the surface parallel to the *n-axis*. It is calculated by *'pretending'* that all other variables ($p$) are constant, and vice versa for $\large \frac{\partial f_n}{\partial p}$.\

Thus the partial derivatives of $\large f_n$ and $\large f_p$ are:

$$\large \frac{\partial f_n}{\partial n} = r-ap,\quad\frac{\partial f_n}{\partial p} = -an$$ $$\large \frac{\partial f_p}{\partial n} = cap,\quad\frac{\partial f_p}{\partial p} = can - \mu$$ Giving us the *Jacobian matrix*:

$$\Large J = \begin{bmatrix} r-ap & -an \\ cap & can - \mu \end{bmatrix}$$

Now, with the added conversion efficiency $\large c$, the equilibrium points are:
$$\Large \text{Prey: } p = \frac{r}{a}, \quad \text{Predator: } n = \frac{\mu}{ca}$$
Lets test the function with these as well.

```{r}
# LV Jacobian matrix
LV_jac <- function(np, P){
  r <- P$r
  a <- P$a
  mu <- P$mu
  c <- P$c
  n <- np[1]
  p <- np[2]
  
  J <- matrix(c(r-a*p, -a*n, c*a*p, c*a*n-mu), nrow = 2, ncol = 2)
  
  return(J)           
}

# Define the parameters
P3 <- list(
  r = 2,
  a = 1,
  mu = 2,
  c = 1
  )

# Define initial population
np <- c(n = 2, p = 1)

# The non-trivial equilibrium points
np_eq <- c(n = P3$mu/(P3$c*P3$a), p = P3$r/P3$a)

# The Jacobian matrix
LV_jac(np, P3)

# The Jacobian matrix at the equilibrium point
LV_jac(np_eq, P3)

```

### d) Calculate the eigenvalues of the Jacobian matrix
Use the `LV_jac()` function to calculate the Jacobian for a set of parameter values. Also calculate the cycle period associated with the complex eigenvalues. Is it a good approximation of the length of the actual cycles?

```{r}
# The Jacobian matrix
J <- LV_jac(np_eq, P3)

eigenval <- eigen(J)

# Compare the eigenvalues to the cycle period
2*pi/Im(eigenval$values[1])

```
