---
title: "Casual Inference"
author: "Andre Bourbonnais"
date: "2023-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}}
library(tidyverse)
library(piecewiseSEM)

```

## Correlation and causation
Correlation does not imply causation. This is a very important concept to understand as it underscores the importance of interpreting results from observational studies.

```{r}

raw_dat <- read.table("../00_DATA/alpineplants.csv", header = TRUE, sep = ",")

glimpse(raw_dat)

```
## Wrightian Path analysis
In its simplest form, a path analysis consists of a series of correlations combined with linear regressions fitted to standardized variables (zero mean, unit variance), thus obtaining *path coefficients*. Before going into technical aspects, a critical point is that before estimating any parameters, causal inference through path analysis or related methods requires formulating a graphical model in the form of a *directed graph* showing the assumed causal (and non-causal) relationships between a set of variables.

As an example, we will consider two different models for how snow depth, minimum winter temperature and soil moisture affect the distribution and abundance of *Carex bigelowii*. In the first model, we will assume independent effects of each predictor, thus building a path model on the form

$$snow \rightarrow Carex.bigelowii$$
$$min.T.winter \rightarrow Carex.bigelowii$$
$$soil.moist \rightarrow Carex.bigelowii$$
An alternative model is that snow cover affects winter temperature and soil moisture, which is turn affects the plant.

$$snow \rightarrow soil.moist$$
$$snow \rightarrow min.T.winter$$

$$min.T.winter \rightarrow Carex.bigelowii$$
$$snow \rightarrow Carex.bigelowii$$

In path analysis, we call the response variables (with arrows coming into them) *endogeneous* variables, and the predictors (with arrows only going out of them) *exogeneous* variables.

The first model can be fitted as a standard multiple regression, while the second model will involve fitting three different component models. Before fitting the models, we remove some `NA`s and *z*-transform all variables (including the response variables).

```{r}
# Remove NA's and scale variables
dat <- raw_dat %>% 
  select(snow, min_T_winter, soil_moist, Carex.bigelowii) %>% 
  na.omit() %>% 
  scale() %>% 
  as.data.frame()

glimpse(dat)
```

We see now that means and standard deviations are zero and one, respectively. 

```{r}
# Means
round(colMeans(dat),2)

```

```{r} 
# Standard deviations
round(apply(dat, 2, sd),2)

```


Lets infer:

- How all the parameters affect *C. bigelowii*.

- How snow affects min_T_winter and soil_moist.

- How min_T_winter and soil_moist affect *C. bigelowii*.
```{r}
# Fit model
m1 <- lm(Carex.bigelowii ~ snow + min_T_winter + soil_moist, data = dat)

# The endogenous variable is on the left side of the formula
# The exogenous variables is on the right side. 
m2a <- lm(min_T_winter ~ snow, data = dat)
m2b <- lm(soil_moist ~ snow, data = dat)

# How does winter and moist affect without snow in the picture.
m2c <- lm(Carex.bigelowii ~ min_T_winter + soil_moist, data = dat)


# First model summary
summary(m1)

```
We notice that they have positive but very small effects on *C. bigelowii* but with weak support. Lets add the estimates, correlations and the unexplained variance.

```{r}
# Snow to min_T_winter
r1 <- cor(dat$snow, dat$min_T_winter, "pairwise") # 0.792
u1 <- sqrt(1 - r1^2) # 0.609

# Snow to soil_moist
r2 <- cor(dat$snow, dat$soil_moist, "pairwise") # 0.443
u2 <- sqrt(1 - r2^2) # 0.896

# min_T_winter to Carex.bigelowii
r3 <- cor(dat$min_T_winter, dat$Carex.bigelowii, "pairwise") # 0.357
u3 <- sqrt(1 - r3^2) # 0.934

# soil_moist to Carex.bigelowii
r4 <- cor(dat$soil_moist, dat$Carex.bigelowii, "pairwise") # 0.197
u4 <- sqrt(1 - r4^2) # 0.980

# snow to Carex.bigelowii
r5 <- cor(dat$snow, dat$Carex.bigelowii, "pairwise") # 0.357
u5 <- sqrt(1 - r5^2) # 0.934


```


## Structrual equation modeling
Structural equation modeling (SEM) is a generalization of path analysis that allows for more complex models, including latent variables. In SEM, we can also fit models with categorical variables, which is not possible in path analysis.


```{r}
# Fit model
m2 = psem(lm(soil_moist~snow, data=dat),
          lm(min_T_winter~snow, data=dat),
          lm(Carex.bigelowii~min_T_winter+soil_moist, data=dat), 
          data=dat)

summary(m2)
plot(m2)
```
















