---
title: "Home_Exam_2"
author: "Andre Bourbonnais"
date: "2023-12-07"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(deSolve)
```

## Protein Production

Protein production (6p) Assume a cell produces a particular protein at a
fixed rate $\large a$. The protein leaks out of the cell membrane at a
rate depending on the protein concentration within the cell. The
dynamics of the protein concentration in the cell can thus be written.

$$ \Large \frac{dP}{dt} = a - bP $$

where $\large P$ is the protein concentration in the cell, $\large a$ is
the production rate of the protein and $\large b$ is the leakage rate of
the protein.

### a) What is the equilibrium concentration of the protein in the cell?

To solve for the equilibrium concentration of the protein in the cell,
we set the derivative of the protein concentration with respect to time
equal to zero and solve for the protein concentration.

$$ \Large \frac{dP}{dt} = a - bP = 0 $$
$$ \Large P = \frac{a}{b} = P^* $$ \### b) Show mathematically that it
is a stable equilibrium. A stable equilibrium is one where the system
will return to the equilibrium point after a small perturbation. An
equilibrium point is stable if: $$ \Large f'(P^*) < 0 $$
$$ \Large f'(P^*) = -b < 0 $$ Thus, the equilibrium point is stable.

### Furthermore,

Assume further that the cell also produces another protein, $\large Q$,
which catalyzes the breakdown of $\large P$. The production of
$\large Q$ is stimulated by the concentration of $\large P$. One can
think of $\large Q$ as a way to regulate the concentration of
$\large P$. The combined dynamics of $\large P$ and $\large Q$ can now
be written:

$$ \Large \frac{dP}{dt} = a - bP - cPQ $$
$$ \Large \frac{dQ}{dt} = rP - bQ $$

where it is assumed that $\large Q$ has the same leaking rate $\large b$
as $\large P$.

### c) Calculate the equilibrium $\large (P^*,Q^*)$ of the system.

First we calculate $\large P^*$ and $\large Q^*$ (i) then we check if
they both are stable equilibrium points (ii).

i)  Calculate $\large P^*$ $$\large 0 = a-bP-cPQ$$
    $$\large bP + cPQ = a$$ $$ \large P^* = \frac{a}{b+cQ}$$

    Calculate $\large Q^*$ $$\large 0 = rP-bQ$$ $$\large bQ = rP$$
    $$\large Q^* = \frac{rP}{b}$$

ii) Check if $\large P^*$ and $\large Q^*$ are stable equilibrium.

### Program in R so that

i)  plots the isoclines in the phase space of the system above.

ii) Simulates the dyamics and plots the resulting trajectory in the same
    phase space.

*"What will the catalyzator (Q) concentration be when the protein
(dP/dt) growth equals zero?"*

$$ \Large \frac{dP}{dt} = a - bP - cPQ = 0 $$
$$ \Large Q = \frac{a-bP}{cP} $$
$$ \Large Q = \frac{a}{cP} - \frac{b}{c} $$

*"What will the protein (P) concentration be when the catalyzator
(dQ/dt) growth equals zero?"*

$$ \Large \frac{dQ}{dt} = rP - bQ = 0 $$ $$ \Large P = \frac{bQ}{r} $$

```{r}


# Define the parameters
params <- list(a = 1, b = 1, c = 1, r = 1)

# 




 





```
