---
title: "Stochastic Systems"
author: "Andre Bourbonnais"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Stochastic systems

A *stochastic system* has an uncertain outcome, it can not be predicted beforehand. One run of a stochastic system is called an *realization* and multiple realizations can be used to *estimate* the outcome of the system. As, we will get a distribution of values for all possible realizations.

```{r, echo = T, message = F, warning = F}
library(tidyverse)
library(ggpubr)
library(viridis)

```

## Exercise: Random walk

Write a function `random_walk()` that simulates $\large N$ number of random walks for $\large t_max$ time-steps, each walk starting at zero. Plot all random walks in the same plot as well a separate histogram for the final position of each random walk.

```{r}
# The function
random_walk <- function(N, t_max){
  
  # Create a dataframe with N columns and t_max rows
  walk_df <- data.frame(matrix(0, nrow = t_max, ncol = N))
  
  # For each walk, generate a random walk
  for(i in 1:N){
    tmp <- numeric(t_max)
    
    # Loop over each time step
    for(j in 2:t_max){
      # Randomly choose a number between 0 and 1
      r <- rnorm(1)
      tmp[j] <- tmp[j-1] + r
    }
    # Update the dataframe
    walk_df[,i] <- tmp
  }
  
  walk_df$time <- 1:t_max
  
  # Return the matrix
  return(walk_df)
}

rwalk <- random_walk(100,1000)

# Make it to long format
rwalk_long <- rwalk %>% 
  pivot_longer(cols = -time, names_to = "walk", values_to = "position")

# Plot the random walks
walks_plot <- rwalk_long %>% 
  ggplot(aes(x = time, y = position, color = walk)) +
  geom_line() +
  theme_bw() +
  scale_color_manual(values = rep("black",ncol(rwalk))) +
  theme(legend.position = "none") +
  labs(x = "Time", y = "Position") +
  ggtitle("Random walks")

# Plot a density plot of the final positions
final_pos_plot <- rwalk_long %>% 
  filter(time == max(time)) %>%
  ggplot(aes(x = position)) +
  geom_density(fill = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Final position", y = "Density") +
  ggtitle("Final position of random walks")

# Plot together
ggarrange(walks_plot, final_pos_plot, ncol = 1, nrow = 2)
```

## Exercise: A queue
Generate a function that simulates a queue. Every time step a new customer is added to the queue with the probability $\large P_{in}$. Furthermore, every time step a customer is finished and leaves the queue with the probability $\large P_{out}$. In the end, plot the number of customers left in the queue over time.


```{r}
# The function
run_queue <- function(Pin, Pout, time_step){
  # Each time step we either add or remove a customer.
  
  queue <- numeric(time_step)
  
  for (ts in 2:time_step){
    # Add a customer
    in_c <- ifelse(runif(1) <= Pin, 1, 0)
    out_c <- ifelse(runif(1) <= Pout, 1, 0)

    # Update the number of customers
    
    queue[ts] <- queue[ts-1]+ in_c + out_c
  }
  
  return(queue)
}

h <- run_queue(0.2, 0.5, 100)

# Plot the queue in ggplot
ggplot(data.frame(h), aes(x = 1:length(h), y = h)) +
  geom_line() +
  theme_bw() +
  labs(x = "Time", y = "Number of customers") +
  ggtitle("Queue")


# plot(h, type = "l", xlab = "Time", ylab = "Number of customers")

```

## Exercise: The Moran process
The Moran process is a model in population genetics, where it models the spread of an allele in a population. The Moran process is a stochastic process, where the population size is constant. The population is divided into two groups, one with the allele and one without. At each time step, one individual is chosen to reproduce, and one individual is chosen to die. The individuals are chosen randomly. The process continues until the population is fixed with the same allele. 

```{r}

moran <- function(N, Na){
  vec <- rep("B", N-Na)
  vec <- c(vec, rep("A", Na))

  for (i in 1:N) {
    
    death_idx <- sample(1:length(vec), 1)
    reprod_idx <- sample(vec[-death_idx], 1)
    vec[death_idx] <- reprod_idx
  }  
  
  return(vec)
}
  
df_moran <- data.frame(matrix(0, ncol = 100, nrow = 10))

for (j in 1:10000) {
  vv <- moran(10,1)
  df_moran[,j] <- vv
}




df_moran_long <- df_moran %>% 
  pivot_longer(cols = everything(), names_to = "time", values_to = "allele")
  
df_A <- df_moran_long %>% 
  group_by(time) %>% 
  mutate(allele.A = sum(allele == "A")) %>% 
  select(time, allele.A)

df111 <- df_A[1:100,]
df111$time <- 1:100
colnames(df111) <- c("iteration", "A")

# Plot the density of the final number of A alleles
df111 %>% 
  ggplot(aes(x = A)) +
  geom_density(fill = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Final number of A alleles", y = "Density") +
  ggtitle("Moran process")






```



 







